Supplementary Materials

### 2022-12-28 manuscript/supplementary.tex
methods stuff

\subsection{Pairwise Diversity Estimates}
\label{supsec:diversity}

@@

As in Section \ref{supsec:diversity}, the number of pairwise differences and
the total number of pairwise comparison at a site are sufficient statistics for
the likelihood. Then, the binomial log-likelihood for the data at site $v$ is,

@@

where $h$ is the bandwidth or window size, $b$ are physical-scale window
indices, $\bar{\pi}(b | \Psi)$ are the average diversity in bin $b$, and
$Y_{\text{S},b}$ and $Y_{\text{S},b}$ are the sums across putatively neutral
sites within a bin. We fit $h$ using out-sample cross validation.

This corresponds to a binomial likelihood for the observed data summarized at
genomic scale $h$. Thus an alternative way to express this model is as,

@@

\subsection{The Background Selection Reduction Factor}

Under the background selection model, the diversity in window $b$ is $\pi(b,
\Psi) = \bar{B}(b, \Psi) \pi_0$. Here, $\bar{B}(b | \Psi)$ is the predicted
reduction in diversity due to BGS in window $b$, given background selection
parameters $\Psi$. In practice, we pre-calculate $\bar{B}(b | \Psi)$ at fixed
sites across the genome, and take the average of the fixed sites $B$ values
within a widow.

In practice, this is site-specific. We can write the
reduction at any neutral site $v$ in the genome as the product of $B$s across
all segments,

\begin{align}
  B(v | \theta) = \exp\left(- \sum_g \int f(\mu(\mathcal{A}(g)), s, S_g) w(s|\mathcal{A}(g)) ds \right)
\end{align}

where $S_g$ is exogenous genomic data about the segment, $S_g = \{L_g, r_g,
\rho(|v-p_g|)\}$, where $L_g$ is the segment's length, $r_g$ is the
recombination rate per basepair in the segment, and $\rho(|v-p_g|)$ is the
recombination distance between the focal site $v$ and the segment position
$p_g$ (we approximate, and use the nearest end position to the neutral site).
Additionally, $w(s|\mathcal{A}(g))$ is the distribution of selection
coefficients for segment $g$, if segment $g$ is a member of annotation class
$\mathcal{A}(g)$.

We can think about the DFE as the conditional distribution of a particular
selection coefficient given a mutation occurs, for a particular annotation
class. The BGS function $f(\cdot)$ only depends on $\mu$ through the
introduction of deleterious alleles with selection coefficient $s$ at rate
$\omega(s|\mathcal{A}(g)) = \mu(\mathcal{A}(g)) w(s | \mathcal{A}(g))$. Thus,
we can write,

\begin{align}
  B(v | \theta) = \exp\left(- \sum_g \int f(\omega(s|\mathcal{A}(g)), s, S_g) ds \right)
\end{align}

which we can discretize as,

\begin{align}
  B(v | \theta) = \exp\left(- \sum_g \sum_s f(\omega(s|\mathcal{A}(g)), s, S_g) \right).
\end{align}

Next, note that there are a finite number of annotation classes, $\mathcal{A}
\to \{a_1, a_2, \ldots, a_k\}$, so we can further partition this as

\begin{align}
  B(v | \theta) = \exp\left(- \sum_{\{g \;:\; \mathcal{A}(g) = a_1\}} \sum_s f(\omega(s| a_1), s, S_g) + \sum_{\{g \;:\; \mathcal{A}(g) = a_2\}} \sum_s f(\omega(s| a_2), s, S_g) + \ldots  \right)
\end{align}

Let us define the $d_\omega \times d_s \times d_g$ multidimensional array
$\mathbf{F}$, and the $d_g \times d_a$ feature classification matrix
$\mathbf{A}$.

\section{Calculation of B Maps}

Each B map is determined by the annotation track of putatively conserved
segments, which includes (1) the segment positions and lengths, (2) the type of
feature of each segment (e.g.  exons, UTRs, phastcons conserved elements,
etc.), as well as (3) the recombination map, and (3) the deleterious selection
coefficient $s(i)$ and mutation rate at which these deleterious enter the
population for feature type $i$.

The B maps are computed at fixed, evenly spaced sites.

>>>>

\subsection{Pairwise Diversity Estimates}
\label{suppsec:diversity}

@@

As in Section \ref{suppsec:diversity}, the number of pairwise differences and
the total number of pairwise comparison at a site are sufficient statistics for
the likelihood. Then, the binomial log-likelihood for the data at site $v$ is,

@@

where $h$ is the bandwidth or window size, $b$ the bin index for windows of
width $h$, $\mathcal{V}_b$ is the set of putatively neutral sites in bin $b$,
$\bar{\pi}(b | \Psi)$ are the average diversity in bin $b$, and
$Y_{\text{S},b}$ and $Y_{\text{S},b}$ are the sums across putatively neutral
sites within a bin. Note that by binning, we sum the pairwise summaries of the
data $\mathbf{Y}$ across sites, so the likelihood across bins is naturally
weighted by the quantity of observed data. We estimate $h$ by fitting the model
for a variety of window sizes $h$ and select the best using out-sample cross
validation, as is standard in non-parametric regression (XXX, Wasserman, p.
70).

This model corresponds to a binomial likelihood for the observed data
summarized at genomic scale $h$. Thus an alternative way to express this model
is as,

@@

\section{The Background Selection Models}

Under the background selection model, the diversity in window $b$ is $\bar{\pi}(b,
\Psi) = \bar{B}(b, \Psi) \pi_0$. Here, $\bar{B}(b | \Psi)$ is the predicted
reduction in diversity due to BGS in window $b$, given background selection
parameters $\Psi$. In practice, we pre-calculate $\bar{B}(x | \Psi)$ at fixed
sites $x$ across the genome, and take the average of the fixed sites $B$ values
within widow $b$ for the average $\bar{B}(b | \Psi)$.

BGS models assume multiplicative fitness effects across sites. Thus, the total
reduction due to segments across the genome is the product of the individual
reductions,

\begin{align}
  \label{suppeqn:b}
  B(x, \Psi) = \prod_g \int_0^1 b(\mu(s, k(g)), s, L_g, r_g, d(x, g)) ds
\end{align}

where, $b(\mu(s, k(g)), s, L_g, r_g, d(x,g))$ is the reduction due to segment
$g$. Here, $\mu(s, k(g))$ is the per-basepair per-genome rate that mutations
with selection coefficient $s$ enter segments with feature class $k(g)$, $L_g$
and $r_g$ are the length in basepairs and recombination rate per basepair of
segment $g$, and the recombination distance between focal site $x$ and segment
$g$ is $d(x, g)$. The functional form of $b$ varies depending on whether the
classic BGS model is used, or our form of Santiago and Caballero's
(\citeyear{Santiago2016-mu}) equations (see Section XXX).

In practice, to fit such a model, we must discretize both $\mu(s, k(g))$ and
$s$. Our models use a fixed $q$-element grid of selection coefficients, and
mutation rates are on a fixed $z$-element grid, which is interpolated over
during ML optimization. The number of annotation features $K$ is variable and
set by the annotation data specified. There are two ways to parameterize the
mutation rates and DFE for all annotation classes in background selection
models. First, for the \emph{free-mutation} parameterization, each annotation
feature class has a free mutation rate parameter for each of these selection
coefficients (as in equation \eqref{suppeqn:b}).

In the free-mutation parameterization, there are $1 + K \times q$ for $\pi_0$
and the free mutation rate parameters, where $K$ is the number of annotation
features. The MLE optimization for this approach is unconstrained, though
$\pi_0$ is bounded to be within $10^{-5} \le \pi_0 \le 10^{-1}$ and mutation
rates are bounded within $10^{-11} \le \mu \le 10^{-7}$. The BGS model requires
a mutation rate for each selection coefficient in the grid, for each annotation
class (here, as an example, CDS, UTRs, and phastcons), which are stored in the
$q \times K$ matrix $\mathbf{M}_\text{F}$,

\begin{align}
  \mathbf{M}_\text{F} = \begin{bmatrix}
    \mu_{10^{-6},\text{CDS}} & \mu_{10^{-6},\text{UTR}} & \mu_{10^{-6},\text{PC}} \\
    \mu_{10^{-5},\text{CDS}} & \mu_{10^{-5},\text{UTR}} & \mu_{10^{-5},\text{PC}} \\
    \vdots & \vdots & \vdots \\
    \mu_{10^{-1},\text{CDS}} & \mu_{10^{-1},\text{UTR}} & \mu_{10^{-1},\text{PC}} \\
  \end{bmatrix}
\end{align}

That is, the distribution of fitness effect (DFE) is implied by
the total mutation rate across selection coefficients for an annotation class.
We can normalize by the total mutation rate estimate to get the estimated DFE,
giving us the DFE weight for annotation class $k$ and selection coefficient
with index $i$,

\begin{align}
  \widehat{w}_{i,k} = \frac{\widehat{\mu}_{i,k}}{\sum_i \widehat{\mu}_{i,k}}.
\end{align}

The second parametrization is the \emph{simplex} parametrization, which has a
single mutation rate across all features, so it is a DFE weight matrix times
the mutation rate $\mu$,

\begin{align}
  \mathbf{M}_\text{S} = \mu \begin{bmatrix}
    w_{10^{-6},\text{CDS}} & w_{10^{-6},\text{UTR}} & w_{10^{-6},\text{PC}} \\
    w_{10^{-5},\text{CDS}} & w_{10^{-5},\text{UTR}} & w_{10^{-5},\text{PC}} \\
    \vdots & \vdots & \vdots \\
    w_{10^{-1},\text{CDS}} & w_{10^{-1},\text{UTR}} & w_{10^{-1},\text{PC}} \\
  \end{bmatrix}
\end{align}

which has $2 + K \times (q-1)$ free parameters, since each column must sum to
one. This imposes the constraint that $\sum_i w_{i,k} = 1$, requiring
constrained MLE optimization.

\subsection{Calculation of the classic BGS B Maps}

The classic BGS model expresses the reduction for a focal site in the middle of
a segment \parencite{Hudson1995-xc,Hudson1994-oh,Nordborg1996-nq}; here we
consider a focal site directly to an arbitrary side of a segment under a
deleterious mutations, and extend the model to handle a focal segment
arbitrarily far from the segment. Using our notation from equation
\eqref{suppeqn:b}, the classic mutation-selection-balance BGS model is,

\begin{align}
  b(\mu, s, L, r, 0) &= \exp \left( - \int_0^L \frac{2\mu(l)}{s (1 + (1-s) r(l)/s)^2} dl \right) 
\end{align}

(c.f. \cite{Hudson1995-xc} equation 5), where $\mu(l)$ is the \emph{haploid}
per-basepair mutation rate at $l$ and $r(l)$ is the recombination fraction
between the focal site (at position 0) and basepair $l$. Assuming constant
per-basepair recombination and mutation rates, $r(l) = r_\text{BP}l$ and
$\mu(l) = \mu$, and

\begin{align}
  b(\mu, s, L, r_\text{BP}, 0) &= \exp \left( - \frac{2\mu L}{s + (1-s) r_\text{BP} L} \right)
\end{align}

(c.f. \cite{Hudson1995-xc} equation 8, where the factor of two difference is
due to the position of the focal site). Often per-region rates are defined $U =
2 \mu L$ and $R = r_\text{BP} L$ are the segment-wide mutation diploid mutation
rates (mutations per diploid segment per generation) and recombination length
(Morgans).

For computational efficiency, we adjust this model by setting $r(l) = d +
r_\text{BP} l$, where $d$ is the recombination fraction between the focal site
$x$ and the start of segment $g$. This allows us to pre-compute the local
effects of the segment, and substitute in $b$ as the focal site changes. This
integral substitution has a closed form,

\begin{align}
  b(\mu, s, L, r_\text{BP}, d) &= \exp \left( \frac{-2\mu L}{(d(1-s) + s)(s + (d + r_\text{BP}L)(1-s))} \right)
\end{align}

and terms can be collected in powers of $d$ and pre-computed for all segments
for the grid of $\mu$ and $s$. Our implementation pre-computes these segment
components, and computes the final $b$ value for an $x$ and $g$ by calculating
the distance $d(x,g) = |M(x) - M(g)|$, where $M(x)$ is the cumulative
recombination map length at position $x$ in Morgans. Since $r(l)$ could vary by
position, segments with differing recombination rates are split into segments
of the same annotation class by their recombination rate. This has no effect on
the calculation due to the multiplicative fitness, but ensures more accurate
estimates of the reduction factor $B$.

% Each B map is determined by the annotation track of putatively conserved
% segments, which includes (1) the segment positions and lengths, (2) the type of
% feature of each segment (e.g.  exons, UTRs, phastcons conserved elements,
% etc.), as well as (3) the recombination map, and (3) the deleterious selection
% coefficient $s(i)$ and mutation rate at which these deleterious enter the
% population for feature type $i$.
